---
title: "IDS — Assignment 1"
author: "Albin Lindqvist (13986236)"
date: "`r Sys.Date()`"
header-includes:
  \usepackage{fvextra}
  \DefineVerbatimEnvironment{Highlighting}{Verbatim}{breaklines,commandchars=\\\{\}}
output:
  pdf_document: default
  html_document: default
---
<style type="text/css">

h1.title {

  text-align: center;
}
h4.author {
  text-align: center;
}
h4.date {
  text-align: center;
}
</style>


# Question 1

Setting up, generating, and saving the data.

```{r}
library(Pareto)
set.seed(100)
Data=data.frame(x.n=rnorm(50000),x.p=rPareto(50000,t=1,alpha=2))
```

## 1.1

Creating the plot, using ggplot.

```{r}
library(ggplot2)
library(gridExtra)
x_n_histo <- ggplot(Data, aes(x=x.n)) + geom_histogram(bins=100, col="black", fill="grey") 
x_n_box <- ggplot(Data, aes(x=x.n)) + geom_boxplot( outlier.color = "black")
grid.arrange(x_n_histo, x_n_box)
```

## 1.2

```{r}
(x_n_mean = mean(Data$x.n))
```

The mean is approximately zero, which is what we would expect from samples of a standard normal distributed variable.

```{r}
(x_n_sd = sd(Data$x.n))
```

The standard deviation is also very close to what we would expect from samples from standard normal distribution.

## 1.3

With the mean we can tell what the average number is. In other words, we can tell in what region the variable should be. So, if we take a sample from the same population, it is most likely we will find the mean, or something that is close to it.

The standard deviation on the other hand is useful to tell how concentrated the variable is, or another way to describe it: it is a measure of the spread of the variable. If we have a high standard deviation we would expect our variable to include a lot of values, and vice versa.

Thus, with this information we can predict, approximately, what the new variable could be. We can see in 1.1 that the samples are symmetric around the mean, which is a reason the mean is good for predicting the new variable. As a matter of fact, we know that the interval $[\mu-2\sigma,\mu+2\sigma]$ should contain around $95\%$ of the samples.

## 1.4

```{r}
x_p_histo <- ggplot(Data, aes(x=x.p)) + geom_histogram(bins=100, col="black", fill="grey")
x_p_box <- ggplot(Data, aes(x=x.p)) + geom_boxplot( outlier.color = "black")
grid.arrange(x_p_histo, x_p_box, x_p_histo + xlim(0,5), x_p_box + xlim(0,5))
cat("Mean of x.p: ", mean(Data$x.p), " and standard deviation of x.p: ", sd(Data$x.p), " the median of x.p: ", median(Data$x.p), ".\n", sep = "")
summary(Data$x.p)
```

We can clearly see that there are some very extreme values of `x.p`, and we need to zoom in a lot to see the box-plot and the histograms properties. We can see that a lot of the values are below $5$, but still quite a lot of values are above it. The standard deviation is around $2.6$ but there are values which are more than 20 times the standard deviation plus the mean, so they are inadequate to describe the data accurately. We can also note that the distribution is not symmetric, there are, for example, no values that are below 1, which is logical because the support of `x.p` is $[1,\infty)$.

If we look at the range of the sample of `x.p` we will see that it is $159.275-1=158.275$, whilst the IQR is $1.992-1.154=0.838$ the difference is staggering, we can thus see that the mean and the standard deviation does bring the characteristics of the distribution.

# Question 2

## 2.1

```{r}
Data2 = read.csv("DataAssignment1.txt", sep="\n", header = FALSE)
Data2_log = log(Data2)
```

As we can see we have some entries that are not a number ('NaN'), this means that our observations contains values of $x<0$, which of course is not defined. So, we will use the following code to find which indexes that are not correct, and also remove them.

```{r}
(idx_na_data2_log = which(is.na(Data2_log)))
cleaned = Data2_log$V1[-idx_na_data2_log]
Data2_log_clean <- data.frame(cleaned)
```

We found out that it was only one value that became 'Nan', so we can relatively safely remove it without altering the data too much. We also confirm that it is indeed gone. We will now plot the log transformed data

```{r}
log_histo <- ggplot(Data2_log_clean, aes(x=cleaned)) + geom_histogram(bins=50, fill="black", col="grey")
log_box <- ggplot(Data2_log_clean, aes(x=cleaned)) + geom_boxplot()
grid.arrange(log_histo, log_box)
```

## 2.2
We will for the questions that follow use the original dataset. We start by plotting the data to see how it is distributed before any transformations.

```{r}
data2_histo <- ggplot(data = Data2, aes(x=V1)) + geom_histogram(bins=100, fill="black", col="grey")
data2_box <- ggplot(data=Data2, aes(x=V1)) + geom_boxplot()
data2_box_iqr <- ggplot(data=Data2, aes(x=V1)) + geom_boxplot(outlier.shape = 0, outlier.stroke = 0) + xlim(median(Data2$V1)-1.5*IQR(Data2$V1), median(Data2$V1)+1.5*IQR(Data2$V1))+xlab("V1 without outliers with IQR-method")

grid.arrange(data2_histo, data2_box, data2_box+xlim(-5,405)+xlab("V1 without largest observation"),data2_box_iqr )
```

It is very obvious we have an outlier, that is very much above any of the others values. We can also note 167 other outliers, according to the IQR-method. However, these outliers are not as significant as the one above $4000$, so we will thus discard that extreme sample as it does not fit the rest of the sample. It is nevertheless worth a short discussion on why it would be.

```{r}
(mean(Data2$V1))
(mean(subset(Data2, V1<1000)$V1))
```

The data is from an insurance company, hence is it is not too unlikely that it was a day they received an extreme claim. Yet, we have to consider the impact this observation have. We can see that if we exclude this one extreme event our average becomes almost half of what it is if we include it, so it is not unreasonable to delete this outlier. We can also not completely rule out if it's a misplaced decimal.

We also have a negative value, which we found in 2.1. It is reasonable to delete this observation because it does not make sense to have a negative claim for an insurance company.

```{r}
# Creating the cleaned dataset
Data2_clean = subset(Data2, V1 >= 0 & V1 < 1000)
```


## 2.3

```{r}
summary(Data2_clean$V1)
summary(subset(Data2_clean, V1 < 2.022+1.5*IQR(Data2_clean$V1))$V1)
```

We can see a clear difference in the median and mean for the whole cleaned data. This is because a lot of our observations are close to 2, but also some observations that are a lot more, that is considered outliers according to the IQR-method (around 17% of the observations). Since the mean is more sensitive to outliers it makes sense that it is much higher than the median, which is a lot less sensitive.

## 2.4
```{r}
rolling_mean <- data.frame(sapply(1:length(Data2_clean$V1), function(i)mean(Data2_clean$V1[1:i])))

rolling_median <- data.frame(sapply(1:length(Data2_clean$V1), function(i)median(Data2_clean$V1[1:i])))

colnames(rolling_mean) <- c("V1")
colnames(rolling_median) <- c("V1")
ggplot() + geom_line(data=rolling_mean, mapping=aes(x=1:length(V1),y=V1,colour="red")) + geom_line(data=rolling_median, mapping=aes(x=1:length(V1),y=V1,colour="blue")) + geom_point(data=Data2_clean, aes(x=1:length(V1),y=V1, colour="green"), alpha=0.2) + labs(x="Day",y="Claims (log10)",title = "All data except extreme outliers", colour="Type") + scale_colour_hue(labels=c("median", "observation", "mean")) + scale_y_log10()

```

These plots shows the clean data (only removing the two extreme outliers), and the rolling mean (red) and median (blue). We use a logarithmic y-axis to be able to represent some of the larger observations. One can note a larger discrepancy between the median and mean and it can be seen that the median stabilizes much quicker than the mean, something we will bear in mind.

```{r}
Data2_no_outliers <- subset(Data2,V1>median(Data2$V1)-1.5*IQR(Data2$V1) & V1<median(Data2$V1) + 1.5*IQR(Data2$V1))

rolling_mean <- data.frame(sapply(1:length(Data2_no_outliers$V1), function(i)mean(Data2_no_outliers$V1[1:i])))

rolling_median <- data.frame(sapply(1:length(Data2_no_outliers$V1), function(i)median(Data2_no_outliers$V1[1:i])))

colnames(rolling_mean) <- c("V1")
colnames(rolling_median) <- c("V1")

ggplot()+geom_line(data=rolling_mean, mapping=aes(x=1:length(V1),y=V1,colour="red")) + geom_line(data=rolling_median, mapping=aes(x=1:length(V1),y=V1,colour="blue")) + geom_point(data=Data2_no_outliers, aes(x=1:length(V1),y=V1, colour="green"),alpha=0.2) + labs(x="Day",y="Claims (log10)",title = "Outliers removed via IQR-method", colour="Type") + scale_colour_hue(labels=c("median", "observation", "mean"))+ scale_y_log10()
```

In this graph, which also have a $\log_{10}$ y-axis, this dataset is a bit smaller and disregards all observations that are outside the interval $[Q_1-1.5\cdot\text{IQR},Q_3+1.5\cdot\text{IQR}]$. One can see that the mean and median, in principle, stabilizes at the same time. As a matter of fact the mean appear to be more stable than the median, something we did not notice in the previous graph.

With these remarks in mind we can start to summarize which method and which measure is preferred. The median appears to be the best when considering both scenarios. When having almost all the data still in use it is still a good measure on the claim amounts. The median tells us that half of it is above, and the other is below this value. However, it does not clearly reflect days where there is a larger claim amount. The dataset that only removed the most extreme outliers should also be preferred as we have no reason to suspect that they are that out of the ordinary and rare.
